---
title: "Assignment 6: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Student Name"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
#This code chunk will tidy your knit PDF files, wrapping long code lines
#For it to work, the "formatR" package needs to be installed

#install.packages('formatR')
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=40), tidy=TRUE)
```

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A06_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup1}
#1
getwd()
library(tidyverse)
library(agricolae)
library(lubridate)
library(cowplot)
library(ggridges)
library(viridis)
library(RColorBrewer)
library(colormap)
library(corrplot)

ChemPhysRaw <- read.csv(
  "./Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv",stringsAsFactors = T)
class(ChemPhysRaw$sampledate)
ChemPhysRaw$sampledate <- as.Date(ChemPhysRaw$sampledate, format= "%m/%d/%y")
class(ChemPhysRaw$sampledate)

#2
iketheme <- theme_bw(base_size = 8) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")
theme_set(iketheme)

```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: Depth has no effect on mean lake temperatures in July across all lakes.
Ha: Depth does have an effect on mean lake temperatures in July across all lakes.


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4
ChemPhysLM <- ChemPhysRaw %>%
  filter(month(sampledate) == 7) %>%
  select(lakename:daynum, depth, temperature_C) %>%
  na.omit()

#5
TempbyDepth <- ggplot(ChemPhysLM, aes(x=depth, y=temperature_C, color = temperature_C))+
  geom_point()+
  scale_color_viridis(option = "magma", direction = 1)+
  geom_smooth(method = lm)+
  ylim(0, 35)+
  xlab("Depth (ft)")+
  ylab("Temperature (C)")
print(TempbyDepth)
```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: The regression line in the plot indicates that as depth increases, temperature decreases. The two variables share an inverse relationship. The distribution of the points shows that we do not have a direct linear relationship between the two variables. 


7. Perform a linear regression to test the relationship and display the results

```{r linear.regression}
#7
Julyregression <- lm(ChemPhysLM$temperature_C ~ ChemPhysLM$depth)
summary(Julyregression)

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: Indicated by the multiple R-squared value, this regression model shows that about 74% of the total variance in temperature is explained by changes in depth. Given that the p-value of the regression is less than 0.05, we reject the null hypothesis, and we accept the hypothesis that depth has an effect on mean lake temperatures across lakes during the month of July. The degrees of freedom (9,726) represent the number of data points that went into the estimation of the relationship between temperature and depth. We can see that the relationship between depth and temperature is negative from the estimate value for ChemPhysRawJuly$depth.


---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9
ChemPhysAIC <- ChemPhysRaw %>%
  filter(month(sampledate) == 7) %>%
  select(lakename:daynum, depth, temperature_C) %>%
  na.omit()


TempAIC <- lm(data = ChemPhysAIC, temperature_C ~ year4 + daynum + depth)
step(TempAIC) #should I be using the AIC and stepwise to test more than these three variables?

#10

MultRegression <- lm(data = ChemPhysAIC, 
                   temperature_C ~ daynum + depth + year4)
summary(MultRegression)

```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: The AIC recommends that we use the "daynum", "depth", and "year4" variables to predict temperature in our multiple regression. Removing each one from the regression resulted in a worse fit. Indicated by the multiple R-squared value, this regression model shows that about 77% of the total variance in temperature is explained by changes in depth, day number, and year. Including these extra explanatory variables imrpoved the variance value by ~3%.


---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
ChemPhysLakesJuly <- ChemPhysRaw %>%
  filter(month(sampledate) == 7) %>%
  select(lakename:daynum, depth, temperature_C) %>%
  na.omit()

#12
# Format ANOVA
Lakes.anova <- aov(data = ChemPhysLakesJuly, temperature_C ~ lakename)
summary(Lakes.anova)

# Format ANOVA as lm
Lakes.anova2 <- lm(data = ChemPhysLakesJuly, temperature_C ~ lakename)
summary(Lakes.anova2)

```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: Based on the fact that the p-values from the ANOVA model and linear regression model are less than 0.05, there is a significant difference for mean water temperature among the lakes. 


14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.
ggplot(ChemPhysLakesJuly, aes(x=depth,y=temperature_C, color=lakename, alpha=0.5))+
  geom_point()+
  geom_smooth(method="lm", se=FALSE)+
  ylim(0,35)+ 
  xlab("Depth (ft)")+
  ylab("Temperature (C)")

```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15
TukeyHSD(Lakes.anova)
Laketest <- HSD.test(Lakes.anova, "lakename", group = TRUE)
Laketest
```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: Using the difference column, Paul Lake and Ward Lake, have similar mean temperatures to Peter Lake. Based on the results of the HSD.test, no lakes have statistically distinct mean temperature from **all** the other lakes.

 
17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: We could use a t-test for this purpose because a t-test 


18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}
ChemPhysCramptonWard <- ChemPhysRaw %>%
  filter(lakename %in% c("Paul Lake", "Peter Lake")) %>%
  filter(month(sampledate) == 7) %>%
  select(lakename:daynum, depth, temperature_C) %>%
  na.omit()

Laketwosample <- t.test(ChemPhysCramptonWard$temperature_C ~ ChemPhysCramptonWard$lakename)
Laketwosample

```

>Answer: Our p-value is 0.013 which is less than 0.05, so we reject the 
null hypothesis that the mean temperature between the two lakes are the same.
The mean temperatures for the lakes are not equal. The mean for Paul Lake is 13.814 while the mean for Peter Lake is 13.316. This result does not match the answer for problem 16, which indicated that Peter Lake and Paul Lake had similar mean temperatures.
