---
title: "Assignment 09: Data Scraping"
author: "Isaac Benaka"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Rename this file `<FirstLast>_A09_DataScraping.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up 
1. Set up your session:

* Check your working directory
* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Set your ggplot theme

```{r, message = FALSE}
#1
getwd()
library(tidyverse)
library(rvest)
library(lubridate)
library(dplyr)
```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2021 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2021>
 
Indicate this website as the as the URL to be scraped. (In other words, read the contents into an `rvest` webpage object.)

```{r set.the.scraping.website}
#2
webpage <- read_html("https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2021")

```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PSWID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Maximum Daily Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to four separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values (represented as strings), with the first value being "27.6400".

```{r scrape.the.data}
#3
water.system.name <- webpage %>% 
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>% 
  html_text()
  
pswid <- webpage %>% 
  html_nodes("td tr:nth-child(1) td:nth-child(5)") %>% 
  html_text()
  
ownership <- webpage %>% 
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>% 
  html_text()
  
max.withdrawals.mgd <- webpage %>% 
  html_nodes("th~ td:nth-child(3) , th~ td+ td:nth-child(6) , th~ td:nth-child(9)") %>%
  html_text()

```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in chronological order. You can overcome this by creating a month column manually assigning values in the order the data are scraped: "Jan", "May", "Sept", "Feb", etc...

5. Create a line plot of the maximum daily withdrawals across the months for 2021

```{r create.a.dataframe.from.scraped.data}
#4
df_withdrawals <- data.frame(
                             "year" = rep(2021),
                             #"Date" = my(paste(Month,"-",Year)),
                             "Water_System_Name" = rep(water.system.name),
                             "PSWID" = rep(pswid),
                             "Ownership" = rep(ownership),
                             "Max_Withdrawals_mgd" = as.numeric(max.withdrawals.mgd))
df_withdrawals$Month <- c("1-1-2021","5-1-2021","9-1-2021","2-1-2021","6-1-2021","10-1-2021","3-1-2021","7-1-2021","11-1-2021","4-1-2021","8-1-2021","12-1-2021")
df_withdrawals$Month <- as.Date(df_withdrawals$Month, format="%m-%d-%Y")
df_withdrawals  <- df_withdrawals %>% arrange(mdy(df_withdrawals$Month)) 
df_withdrawals <- df_withdrawals %>% rename(Date=Month)


#5
ggplot(df_withdrawals,aes(x=Date,y=Max_Withdrawals_mgd))+
  geom_line()+
  ylab("Maximum Withdrawals (MGD)")


```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function using your code above that can scrape data for any PWSID and year for which the NC DEQ has data. **Be sure to modify the code to reflect the year and site (pwsid) scraped**.

```{r construct.a.scraping.function}
#6.
scrape.it <- function(PSW,the_year){
theURL <- read_html(paste0("https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=",PSW,"&year=",the_year,"/"))

water.system.tag <- "div+ table tr:nth-child(1) td:nth-child(2)"
PSW.tag <- "td tr:nth-child(1) td:nth-child(5)"
ownership.tag <- "div+ table tr:nth-child(2) td:nth-child(4)"
max.withdrawals.mgd.tag <- "th~ td:nth-child(3) , th~ td+ td:nth-child(6) , th~ td:nth-child(9)"

Water.system.name <- theURL %>% 
  html_nodes(water.system.tag) %>% 
  html_text()
  
PSW <- webpage %>% 
  html_nodes(PSW.tag) %>% 
  html_text()
  
Ownership <- webpage %>% 
  html_nodes(ownership.tag) %>% 
  html_text()
  
Max.withdrawals.mgd <- webpage %>% 
  html_nodes(max.withdrawals.mgd.tag) %>%
  html_text()

df_withdrawals_scrape <- data.frame(
  Year=rep(the_year),
  Date=c("1-1-2021","5-1-2021","9-1-2021","2-1-2021","6-1-2021","10-1-2021","3-1-2021","7-1-2021","11-1-2021","4-1-2021","8-1-2021","12-1-2021"),
  PSWID = rep(PSW),
  Ownership = rep(Ownership),
  Max_Withdrawals_mgd = as.numeric(Max.withdrawals.mgd)
  
)
}

```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2015

```{r fetch.and.plot.Durham.2015.data}
#7
withdrawals.2015 <- scrape.it("03-32-010" , "2015")
return(withdrawals.2015)
```

8. Use the function above to extract data for Asheville (PWSID = 01-11-010) in 2015. Combine this data with the Durham data collected above and create a plot that compares Asheville's to Durham's water withdrawals.

```{r fetch.and.plot.Asheville.2015.data}
#8


```


9. Use the code & function you created above to plot Asheville's max daily withdrawal by months for the years 2010 thru 2019.Add a smoothed line to the plot.

>TIP: See Section 3.2 in the "09_Data_Scraping.Rmd" where we apply "map2()" to iteratively run a function over two inputs. Pipe the output of the map2() function to `bindrows()` to combine the dataframes into a single one. 

```{r fetch.and.plot.Asheville.multiyear.data}
#9


```

>Question: Just by looking at the plot (i.e. not running statistics), does Asheville have a trend in water usage over time?